---
title: "tuning_the_data"
author: "Adam Zabell"
date: "12/21/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
workingDir <- "/Users/azabell/Dropbox/R/mockData MSACL2018"
useNewerPackage <- FALSE # If running R 3.3.0 or newer, use openxlsx
if(useNewerPackage) {
 #install.packages("openxlsx")
 library(openxlsx)
} else {
 #install.packages("xlsx")
 library(xlsx)
}
```

# Making database repairs 
After construction of the data files, several gaps were discovered. Most notably, there were five days of batches where a timestamp and a reviewer were not added. Other edits will be applied from this file as their need is discovered.

## 1. Missing timestamps
```{r missing_timestamp}
add_timestamps_postproduction <- function(batch.df,batchNames,dateCollected="2017-01-01") {
  # data collected 8pm-2am, reviewStart 9am-2pm, avg batch review time is 1 hr
  dateAs.posixlt <- strptime(paste(dateCollected,"20:00:00"),"%F %X") 
  for(b in batchNames) {
    doneTime <- dateAs.posixlt + round(runif(1,0,21600),0)
    startTime <- dateAs.posixlt + round(runif(1,46800,64800),0)
    endTime <- startTime + round(rnorm(1,3600,900),0)
    if(endTime > startTime) {endTime <- startTime + round(runif(1,900,2700),0)}
    batch.df$batchCollectedTimestamp[batch.df$batchName == b] <- as.character(doneTime)
    batch.df$reviewStartTimestamp[batch.df$batchName == b] <- as.character(startTime)
    batch.df$reviewCompleteTimestamp[batch.df$batchName == b] <- as.character(endTime)
  }
  return(batch.df)
}

allRdata <- list.files(path=workingDir,pattern="Rdata$",full.names=TRUE)
for(i in 1:12) {
  load(allRdata[i])
  didNotTimestamp <- which(is.na(batchLevel$batchCollectedTimestamp))
  if(length(didNotTimestamp) == 0) {next}
  batchNames <- unique(batchLevel$batchName[didNotTimestamp])
  justBefore <- batchLevel[didNotTimestamp[1]-1,]
  noon <- paste(as.Date(justBefore$batchCollectedTimestamp),"12:00:00 GMT")
  isAfterNoon <- justBefore$batchCollectedTimestamp > noon
  todaysDate <- as.Date(justBefore$batchCollectedTimestamp)
  if(isAfterNoon) {todaysDate <- todaysDate + 1}
  batchLevel <- add_timestamps_postproduction(batchLevel,batchNames,dateCollected=as.character(todaysDate))
  firstDate <- as.Date(min(batchLevel$batchCollectedTimestamp))
  save(list=c("batchLevel","compoundLevel","peakLevel"),
     file=file.path(workingDir,paste0(as.character(firstDate),".Rdata")) )
  excelFile <- paste0(as.character(firstDate),".xlsx")
  write.xlsx2(batchLevel, file=file.path(workingDir,excelFile), row.names=FALSE )
}
```

## 2. Missing reviewers
This looks much like the prior section, with an extra function (hidden in the report but present in the script) to make sure the course lessons don't get broken.
```{r missing_reviewer_adjusting_function, echo=FALSE}
adjust_reviewer_index <- function(randomNameAssignments,i) {
  if(i==2 | i==3) {
    #no Zachary
    randomNameAssignments[randomNameAssignments==6] <- 2
  }
  if(i==4) {
    #no Walter
    randomNameAssignments[randomNameAssignments==4] <- 6
  }
  if(i==5) {
    #no Dave
    randomNameAssignments[randomNameAssignments==4] <- 3
    #no Walter
    randomNameAssignments[randomNameAssignments==5] <- 6
  }
  return(randomNameAssignments)
}
```
```{r missing_reviewer}
add_reviewer_postproduction <- function(batch.df,batchNames,reviewerName) {
  for(b in batchNames) {
    batch.df$reviewerName[batch.df$batchName == b] <- reviewerName
  }
  return(batch.df)
}

allReviewerNames <- list(Alice=c(1:12),
                         Brad=c(1:12),
                         Charles=c(1:12),
                         Dave=c(5:12),
                         Elizabeth=c(6:12),
                         Frank=c(6:12),
                         Walter=c(4:5),
                         Xavier=c(1:3),
                         Yolanda=c(1:5),
                         Zachary=c(1:4))
allRdata <- list.files(path=workingDir,pattern="Rdata$",full.names=TRUE)
for(i in 1:12) {
  load(allRdata[i])
  didNotAddReviewer <- which(batchLevel$reviewerName=="none")
  if(length(didNotAddReviewer) == 0) {next}
  batchNames <- unique(batchLevel$batchName[didNotAddReviewer])
  currentReviewers <- names(allReviewerNames)[sapply(allReviewerNames,function(x){i%in%x})]
  randomNameAssignments <- ceiling(runif(length(batchNames))*6)
  randomNameAssignments <- adjust_reviewer_index(randomNameAssignments,i)
  for(idx in 1:6) {
    batchSubset <- which(randomNameAssignments==idx)
    if(length(batchSubset) == 0) {next}
    batchLevel <- add_reviewer_postproduction(batchLevel,batchNames[batchSubset],currentReviewers[idx]) 
  }
  firstDate <- as.Date(min(batchLevel$batchCollectedTimestamp))
  save(list=c("batchLevel","compoundLevel","peakLevel"),
     file=file.path(workingDir,paste0(as.character(firstDate),".Rdata")) )
  excelFile <- paste0(as.character(firstDate),".xlsx")
  write.xlsx2(batchLevel, file=file.path(workingDir,excelFile), row.names=FALSE )
}
```

## 3. Rewrite the Year
```{r rewrite_year}
  allRdata <- list.files(path=workingDir,pattern="Rdata$",full.names=TRUE)
  bigBatchLevel <- list()
  bigCompoundLevel <- list()
  bigPeakLevel <- list()
  for(r in 1:12) {
    load(allRdata[r])
    m <- switch(r,
                "January","February","March","April","May","June",
                "July","August","September","October","November","December")
    bigBatchLevel[[m]] <- batchLevel
    bigCompoundLevel[[m]] <- compoundLevel
    bigPeakLevel[[m]] <- peakLevel
  }
  batchYear <- do.call(rbind,lapply(bigBatchLevel, function(x) {return(x)}))
  compoundYear <- do.call(rbind,lapply(bigCompoundLevel, function(x) {return(x)}))
  peakYear <- do.call(rbind,lapply(bigPeakLevel, function(x) {return(x)}))
  save(list=c("batchYear","compoundYear","peakYear"),
     file=file.path(workingDir,"wholeYear.Rdata") )
```