---
title: 'Lesson 5: Blending data from multiple files and sources'
author: "Randall Julian, Adam Zabell"
date: "9/3/2018"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Joining Relational Data

The database example for this class has three different tibbles: one for batch-level information (calibration R^2^, instrument name); one for sample-level information (sample type, calculated concentration); and one for peak-level information (quant peak area, modification flag). Accessing the relationships across these three sources -- reporting the quant and qual peak area of only the qc samples, for example -- requires the tools of relational data. In the tidyverse, these tools are part of the **dplyr** package and involve three 'families of verbs' called *mutating joins*, *filtering joins*, and *set operations*, which in turn expect a unique key in order to correctly correlate the data. To begin, read in the batch, sample, and peak data from the month of January. For simplicity, we will reduce size of our working examples to only those rows of data associated with one of two batches.  
```{r one_batch_three_tables, message=FALSE, eval = FALSE}
january_batches <- read_csv("data/2017-01-06_b.csv")
january_samples <- read_csv("data/2017-01-06_s.csv")
january_peaks <- read_csv("data/2017-01-06_p.csv")

byBatch <- january_batches[january_batches$batchName=="b802253" |
                           january_batches$batchName=="b252474",]

bySample <- january_samples[january_samples$batchName=="b802253" |
                            january_samples$batchName=="b252474",]

byPeak <- january_peaks[january_peaks$batchName=="b802253" | 
                        january_peaks$batchName=="b252474",]
```


## Blending Data

### Simple *bind_rows* and *bind_cols*
Sometimes, you need to combine data stored in more than one file. For example, managing the QC deviations across twelve separate months of reports. To do this in R, you can read each file and then merge them together either by row, or by column. The idea behind *tidy data* is that each column is a variable, each row is an observation, and each element is a value. If you know that your data sources have the same shape (same variables and same observations), you can safely combine them with an `bind_rows` to append the second source of data at the end of the first.  
```{r bind_rows, eval = FALSE}
january_samples <- read_csv("data/2017-01-06_s.csv")
as.data.frame(january_samples[187195:187200,])
february_samples <- read_csv("data/2017-02-06_s.csv")
as.data.frame(february_samples[1:5,])
twoMonths <- bind_rows(january_samples,february_samples)
```

Notice the continuation from the last rows of `january` to the first rows of `february` and that the number of rows in the combined data.frame `twoMonths` is the sum of the first two months of sample-level data.  
```{r jan_plus_feb, eval = FALSE}
twoMonths[187195:187204,]
c(nrow(january_samples), nrow(february_samples), nrow(twoMonths))
```

As long as the two tibbles have the same number of columns and the same column names, the `bind_rows` command will correctly associate the data using the column order from the first variable. And if they aren't the same, you get an error that tells you what is wrong. That makes `bind_rows` useful but remeber to the data are clean before you use.

**Exercise 1: Using `bind_rows`**
Try to use `bind_rows()` to combine all of the data from Perform an rbind between February and each of these three tibbles containing January data. Do any of them work? What does the data look like? What error messages do you get?  

```{r rbind_ex, eval = FALSE}
bind_rows(january_peaks,   february_samples)
bind_rows(january_batches, february_samples)
bind_rows(january_samples, february_samples) # the only January data.frame with matching columns
```
**End of Exercise**

There is an related command called `bind_cols` which will append columns to a tibble, but it also requires very clean data. This command will not check to make sure the order of values are correct between the two things being bound.
```{r cbind, eval = FALSE}
incomplete_data <- tibble(sampleName="123456",
                          compoundName=c("morphine","hydromorphone",
                                         "codeine","hydrocodone"),
                          concentration=c(34,35,44,45))

additional_columns <- tibble(expectedConcentration=c(20,30,40,40),
                             sampleType="standard")

desired_bind   <- bind_cols(incomplete_data,additional_columns)
```

### Binding using relationships between data objects

Using *dplyr* there is another way of binding data which does not require the items being combined to be identical in shape. It does require adopting a relational database approach to the design of your data structures. This is, at the core, the primary idea behind *tidy* data.

### Primary and foreign keys

A key is the variable in a tibble -- or combination of variables in a tibble -- that uniquely defines every row. In our data, `batchName` is present in each tibble but is insufficient to define a specific row. As it turns out for this data set, no single column operates as a key. We can build a key by combinging two (or three) columns. Here is how to combine values which are not unique to an individual observation in order to create a key which is unique to each observation.

```{r uniquify, eval = FALSE}
byBatch$keyB <- paste(byBatch$batchName, byBatch$compoundName, sep=":")
#bySample$keyS <- paste(bySample$sampleName, bySample$compoundName, sep=":")
#byPeak$keyP <- paste(byPeak$sampleName, byPeak$compoundName, 
#                     byPeak$chromatogramName, sep=":")
```

This creates what is call a *primary key*, which is the unique identifier for each observation in a specific tibble. A *foreign key* is the same thing, only it uniqely identifies an observation in another tibble. The `left_join` command joins two tibbles based on matching the *primary key* in the first tibble with the *foreign key* in the second tibble.    

```{r byPeak_foreign_key, eval = FALSE}
bySample$keyB <- paste(bySample$batchName, bySample$compoundName, sep=":")
combined <- left_join(bySample, byBatch, by="keyB")
```

### Set operations *union*, *intersect*, and *setdiff*

Relational databases operate using operations from set theory. Joining two datasets by matching a common identifier is the most common application, however *dplyr* supports many more set operations which are useful in wrangling complex data. Unions, intersections, and the differences between dataset are often needed to get to the data you want. 

These three commands will return a vector which is the unduplicated combination of the two input vectors. `union(A,B)` includes all the values found in both A and B. `intersect(A,B)` returns only those values found in both A and B. `setdiff(A,B)` is order dependent, and returns the values of the first vector which are not also in the second vector. 
```{r set_operations, eval = FALSE}
A <- rep(seq(1, 10), 2)
B <- seq(2, 20, 2)
union(A, B)
intersect(A, B)
setdiff(A, B)
setdiff(B, A)
```

These commands are good for checking matches between two vectors, and we can use them to rebuild the `byPeak$keyB` foreign key without the risk of incorrect naming. First, let's reset `byPeak`. 
```{r byPeak_reset, eval = FALSE}

byPeak <- january_peaks %>% 
  filter(batchName %in% c("b802253","b252474")) %>%
  unite(keyP, sampleName, compoundName, chromatogramName, sep=":", remove=FALSE)
```

Now let's construct our `byPeak$keyB` foreign key by creating and then using a new variable called *analyte*, taking advantage of set operations.

```{r better_foreign_key, eval = FALSE}
allNames <- unique(byPeak$compoundName)
byPeak$analyte <- NA
for (name in allNames[1:6]) {
  compoundPairIdx <- grep(name, allNames)
  theCompound <- intersect(allNames[compoundPairIdx], name)
  theInternalStandard <- setdiff(allNames[compoundPairIdx], name)
  byPeak$analyte[byPeak$compoundName == theInternalStandard] <- theCompound
  byPeak$analyte[byPeak$compoundName == theCompound] <- theCompound
}
```

## Mutating join to add columns 
Mutating joins operate in much the same way as the set operations, but on data.frames instead of vectors, and with one critical difference: repeated values are retained. We took advantage of this earlier when using the left_join command, so that the `byBatch$keyB` got repeated for both the Quant and the Qual peak entries in `byPeak`. Having built the `byBatch` primary key, and correctly included it as a foreign key in `byPeak`, correctly joining them into a single data.frame is straightforward. 
```{r using_primary_and_foreign_keys, eval = FALSE}
byPeakWide <- left_join(byPeak,byBatch)
```

There are four kinds of mutating joins, differing in how the rows of the source data.frames are treated. In each case, the matching columns are identified automatically by column name and only one is kept, with row order remaining consistent with the principle (usually the left) source. All non-matching columns are returned, and which rows are returned depends on the type of join. An *inner_join(A,B)* only returns rows from A which have a column match in B. The *full_join(A,B)* returns every row of both A and B, using an NA in those columns which don't have a match. The *left_join(A,B)* returns every row of A, and either the matching value from B or an NA for columns with don't have a match. Finally, the *right_join(A,B)* returns every row of B, keeping the order of B, with either the matching value from columns in A or an NA for columns with no match.

At first it may be confusing that some joins result in making duplicates of rows from one tibble. However, it is extremely handy when you are using one dataset to label another.
```{r duplication_ex, eval = FALSE}
goodDuplication <- inner_join(
  x = bySample[, c(1:3, 7)],
  y = byBatch[, c(1:6)],
  by = c("batchName", "compoundName")
)
```

## Filtering join to check the overlap
We created the `byBatch$keyB` explicitly by looking directly at the data. The compound naming scheme in `byPeak` is more complicated in that the internal standard isn't identified in `byBatch` or `bySample`, so we fixed this using a new column *analyte*. 

Sometimes it is useful to look for these complications using the semi_join and anti_join commands. The *semi_join(A,B)* returns all rows of A where there is a match from B, but keeps only the columns of A, and does not duplicate a row if there are multiple matches. The *anti_join(A,B)* is the inverse, returning all rows from A where there is no match from B. We still want to create the 'analyte' column for clarity, so one approach would be:
```{r filtering_joins, eval = FALSE}
byBatch <- january_batches %>% # reset, no keyB 
  filter(batchName %in% c("b802253","b252474"))
byPeak <- january_peaks %>% # reset, no keyB or keyP 
  filter(batchName %in% c("b802253","b252474")) %>%
  mutate(analyte=compoundName)

unique(byPeak$analyte) # notice the similar naming scheme?
byPeak$analyte <- sub("-.*$", "", byPeak$analyte) # notice how we used that similarity?

noMatch <- anti_join(byPeak,byBatch)
noMatch <- anti_join(byPeak,byBatch,by=c("batchName","analyte"="compoundName"))

justMatch <- semi_join(byPeak,byBatch,by=c("batchName","analyte"="compoundName"))
```

**Exercise 2: Join the batch and peak data**  
Start from the reset tibbles built in the prior code chunk, so the `keyB` and `keyP` variables are not present.

```{r join_batch_peak, eval = FALSE}
byPeak$analyte <- sub("-.*$", "", byPeak$analyte) # ensure the variable has been modified
exerciseThree <- left_join(byBatch, byPeak,by=c("batchName","compoundName"="analyte") )
```

## Summary
* `rbind` and `cbind` add rows (or columns) to an existing data.frame
* `union`, `intersect`, and `setdiff` return a combination of two vectors
* Relational data merges two data.frames on the common columns, called keys
    + A primary key is a unique identifier for every row in a data.frame (the presence of `keyB` in `byBatch`)
    + A foreign key is a unique identifier for another data.frame (the presence of `keyB` in `byPeak`)
* `inner_join`, `full_join`, `left_join`, and `right_join` are mutating joins which add columns
* `semi_join` and `anti_join` are filtering joins which check for overlap