---
title: 'Lesson 6: Blending data from multiple files and sources'
output:
  html_document: default
---

```{r setup_6, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidylog, warn.conflicts = FALSE)
library(readxl)
library(janitor)
```

## Joining Relational Data

The database example for this class has three different tibbles: one for batch-level information (calibration R^2^, instrument name); one for sample-level information (sample type, calculated concentration); and one for peak-level information (quant peak area, modification flag). Accessing the relationships across these three sources -- reporting the quant and qual peak area of only the qc samples in specific batches by instrument, for example -- requires the tools of relational data. In the tidyverse, these tools are part of the **dplyr** package and involve three 'families of verbs' called *mutating joins*, *filtering joins*, and *set operations*, which in turn expect a unique key in order to correctly correlate the data. To begin, read in the batch, sample, and peak data from the month of January. For simplicity, we will reduce size of our working examples to only those rows of data associated with one of two batches.  

```{r, message=FALSE}
january_batches <- read_csv("data/2017-01-06_b.csv") %>%
  clean_names() #use help to check out what this does
january_samples <- read_csv("data/2017-01-06_s.csv") %>%
  clean_names()
january_peaks <- read_csv("data/2017-01-06_p.csv") %>%
  clean_names()
select_batches <- january_batches %>%
  filter(batch_name %in% c("b802253", "b252474"))
select_samples <- january_samples %>%
  filter(batch_name %in% c("b802253", "b252474"))
select_peaks <- january_peaks %>%
  filter(batch_name %in% c("b802253", "b252474"))
```


## Blending Data

### Simple addition of rows and columns

Sometimes, you need to combine data stored in more than one file. For example, managing the QC deviations across twelve separate months of reports. To do this in R, you can read each file and then merge them together either by row, or by column. The idea behind *tidy data* is that each column is a variable, each row is an observation, and each element is a value. If you know that your data sources have the same shape (same variables and same observations), you can safely combine them with an `bind_rows` to append the second source of data at the end of the first. 

```{r}
january_samples <- read_csv("data/2017-01-06_s.csv") %>%
  clean_names()

february_samples <- read_csv("data/2017-02-06_s.csv") %>%
  clean_names()

two_months <- bind_rows(january_samples, february_samples)

```

Notice the continuation from the last rows of january to the first rows of february and that the number of rows in the combined data frame two_months is the sum of the first two months of sample-level data. Another way to see this is to add a column name for the ".id" argment in the `bind_rows` call.

```{r}
two_months_id <- bind_rows(january_samples, february_samples, .id = "month") #.id is the dataframe identifier argument - will add new column id with indicator of which dataframe a row comes from 1 = january_samples; 2 = february_samples.
```


```{r}
two_months[187195:187204,]
c(nrow(january_samples), nrow(february_samples), nrow(two_months))

two_months_id[187195:187204,]

```

As long as the two tibbles have the same number of columns and the same column names, the `bind_rows` command will correctly associate the data using the column order from the first variable. And if they aren't the same, you get an error that tells you what is wrong. That makes `bind_rows` useful but remember to make sure the data are clean before you use this function.  

**Exercise 1**

Try to use `bind_rows()` to combine all of the sample data from February and each of the three tibbles containing January data. Do any of them work? What does the data look like?  

First try binding a peaks file with the February samples file:


```{r, echo = FALSE, eval = FALSE}
# complete this code
peaks_samples <- bind_rows()
summary(peaks_samples)
```


```{r}
peaks_samples <- bind_rows(january_peaks, february_samples)
summary(peaks_samples) # bind_rows completes, but adds all columns in data set with NAs
```

Observe the number of columns and visualize the new data frame directly.

Next try the batches and samples file:

```{r, echo = FALSE, eval = FALSE}
#complete this code
batches_samples <- bind_rows()
summary(batches_samples)
```

```{r}
batches_samples <- bind_rows(january_batches,   february_samples)
summary(batches_samples) # bind_rows completes, but adds all columns in data set with NAs
```

Now bind both samples files:

```{r, echo = FALSE, eval = FALSE}
#complete this code
batches_samples <- bind_rows()
summary(batches_samples)
```

```{r}
samples_samples <- bind_rows(january_samples, february_samples) # no extra columns
summary(samples_samples)
```
**End Exercise**

There is an related command called `bind_cols` which will append columns to a tibble, but it also requires very clean data. This command will not check to make sure the order of values are correct between the two things being bound.

```{r}
incomplete_data <- tibble(sampleName="123456",
                          compoundName=c("morphine","hydromorphone",
                                         "codeine","hydrocodone"),
                          concentration=c(34,35,44,45))

additional_columns <- tibble(expectedConcentration=c(20,30,40,40),
                             sampleType="standard")

desired_bind   <- bind_cols(incomplete_data, additional_columns)
head(desired_bind)
```

### Binding using relationships between data objects

Using *dplyr* there is another way of binding data which does not require the items being combined to be identical in shape. It does require adopting a relational database approach to the design of your data structures. This is, at the core, the primary idea behind *tidy* data.

### Primary and foreign keys

A key is the variable in a tibble -- or combination of variables in a tibble -- that uniquely defines every row. In our data, `batch_name` is present in each tibble but is insufficient to define a specific row. If we want to join data from different tables and ensure it matches to the correct row, we need a key. As it turns out for this data set, no single column operates as a key. We can build a key by combining two (or three) columns. Here is how to combine values which are not unique to an individual observation in order to create a key which is unique to each observation. We create the key for the `select_peaks` data using a dplyr alternative function to `paste()` (base R) called `unite()`. This function takes the data as the first argument (piped in this examples), and then will put together specified columns using a separator you specify. If you don't want to remove the variables used to construct the key, you add the "remove = FALSE" argument.

```{r}
select_batches <- select_batches %>%
  unite(keyB, c(batch_name, compound_name), sep=":", remove = FALSE)

```

This creates what is called a *primary key*, which is the unique identifier for each observation in a specific tibble. A *foreign key* is the same thing, only it uniquely identifies an observation in another tibble. The `left_join` command joins two tibbles based on matching the *primary key* in the first tibble with the *foreign key* in the second tibble.  

```{r}
#create keys in peaks and samples tables
select_peaks <- select_peaks %>%
  unite(keyB, c(batch_name, compound_name), sep=":", remove = FALSE)

select_samples <- select_samples %>%
  unite(keyB, c(batch_name, compound_name), sep=":", remove = FALSE)

#join by key
combined <- left_join(select_samples, select_batches, by= "keyB")

```

## Mutating join to add columns

Mutating joins operate in much the same way as the set operations (`union()`, `intersect()`, `setdiff()`), but on data frames instead of vectors, and with one critical difference: repeated values are retained. We took advantage of this earlier when using the left_join command, so that the `select_batches$keyB` got repeated for both the Quant and the Qual peak entries in `select_peaks`. Having built the `select_batches` primary key, and correctly included it as a foreign key in `select_peaks`, correctly joining them into a single data frame is straightforward.   

```{r}

select_peaksWide <- left_join(select_peaks,select_batches)

```

There are four kinds of mutating joins, differing in how the rows of the source data frames are treated. In each case, the matching columns are identified automatically by column name and only one is kept, with row order remaining consistent with the principle (usually the left) source. All non-matching columns are returned, and which rows are returned depends on the type of join. An *inner_join(A,B)* only returns rows from A which have a column match in B. The *full_join(A,B)* returns every row of both A and B, using an NA in those columns which don't have a match. The *left_join(A,B)* returns every row of A, and either the matching value from B or an NA for columns with don't have a match. Finally, the *right_join(A,B)* returns every row of B, keeping the order of B, with either the matching value from columns in A or an NA for columns with no match.


**Exercise 2:** 

Join the data from the `select_samples` dataset with the data from `select_peaksWide`. Try using a `left_join()` and a `right_join()` to see the difference. We want to join the peaks and batches data to the samples in a way that eliminates the rows with internal standard information -- so retain data that is in `select_samples`.  

```{r, echo = FALSE, eval = FALSE}
#complete this code
select_all <- left_join() 

#alternative using right_join
select_allR <- right_join() 
```

```{r}
#data from all 3 tables joined to retain data based on select_samples
select_all <- left_join(select_samples, select_peaksWide) #notice which is first

#alternative
select_allR <- right_join(select_peaksWide, select_samples) #notice reverse in right_join
```


**End Exercise**

## Back to our problem

We started out with the intention to combine data from three tables so we could report qualifier and quantifier peak areas of QC samples from select batches, noting which instrument was used. We now know how to create a dataset we need for this analysis.  

**Exercise 3**

(1) Join `january_peaks`, `january_batches`, and `january_samples`. *Hint: first create a key in each table. Join to retain rows in samples table.*  

```{r, echo = FALSE, eval = FALSE}
#create keys in each table



#join tables to retain rows from samples table

```


(2) Filter this joined dataset for QCs and group by instrument, compound name, expected concentration, and chromatogram.  

Summarize the grouped data to find the mean, sd, and cv of peak areas and the number of qcs for a given condition. 

```{r, echo = FALSE, eval = FALSE}



```


(3) Create a graphic from the joined dataset showing boxplots of peak areas for the qualifier and quantifier peaks of each compound at each expected QC concentration, colored by instrument.

```{r, echo = FALSE, eval = FALSE}


```


```{r}
#create primary keys in each table
january_peaks <- january_peaks %>%
  unite(keyB, c(batch_name, compound_name), sep=":", remove = FALSE)

january_batches <- january_batches %>%
  unite(keyB, c(batch_name, compound_name), sep=":", remove = FALSE)

january_samples <- january_samples %>%
  unite(keyB, c(batch_name, compound_name), sep=":", remove = FALSE)

#join peaks with batches and then with samples to retain data from samples
jan_peaks_batches <- left_join(january_peaks,january_batches)
jan_samples_peaks_batches <- right_join(jan_peaks_batches, january_samples)
```

```{r}
#group by conditions and summarize
qc_table <- jan_samples_peaks_batches %>% 
  filter(sample_type == "qc") %>% 
  group_by(instrument_name, compound_name, expected_concentration, chromatogram_name) %>%
  summarize(mean_area = mean(peak_area),
            sd_area = sd(peak_area),
            cv_area = sd_area/mean_area,
            n = n()) 
```

```{r}
ggplot(filter(jan_samples_peaks_batches, sample_type == "qc")) +
  geom_boxplot(aes(x = compound_name, y = peak_area, color = instrument_name)) +
  facet_grid(as.factor(expected_concentration)~chromatogram_name, scales = "free") +
  labs(x = " ", y = "Peak Area", color = "Instrument") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        panel.spacing.x = unit(3.6, "mm"))
```

Bonus!
Create a boxplot showing QC concentration by compound and instrument for each expected concentration.

```{r, echo = FALSE, eval = FALSE}

```


```{r}
ggplot(filter(jan_samples_peaks_batches, sample_type == "qc", chromatogram_name == "quant")) +
  geom_boxplot(aes(x = compound_name, y = concentration, color = instrument_name)) +
  facet_grid(as.factor(expected_concentration)~., scales = "free") +
  labs(x = " ", y = "Concentration", color = "Instrument") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        panel.spacing.x = unit(3.6, "mm"))
```


## Summary

- `rbind` and `cbind` add rows (or columns) to an existing data frame
- Relational data merges two data frames on the common columns, called keys
  - A primary key is a unique identifier for every row in a data frame (the presence of `keyB` in `select_batches`)
  - A foreign key is a unique identifier for another data frame (the presence of `keyB` in `select_peaks`)
- `inner_join`, `full_join`, `left_join`, and `right_join` are mutating joins which add columns of one table to another