---
title: 'Lesson 8: Exploratory Data Analysis'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(janitor)
library(lubridate)
```

## Overview of lesson activities

In this lesson we will gain more experience with some of the tools we have discussed throughout this course and ask you to dive into a new data set to answer a variety of questions. For many of the questions we will ask, there is no right or wrong way to answer the question. However, this is an opportunity to use new functions you have learned so far in this course. Our answers to the questions will primarily use tidyverse functions, but regardless of how you answer questions, you are looking for output of code to be the same.

## Introduction to data set

The data set for this lesson is derived from orders for clinical laboratory tests in an electronic health record system in a set of outpatient clinics. The orders were deidentified and time-shifted (and approved for use as a teaching resource). There are two files:
- "orders_data_set.xlsx" represents the data as one row per order and includes the bulk of the details
- "order_details.csv" maintains the one row per order structure and include ancillary information about how a test was ordered

There are some column pairs with very similar names: one variable is a code ("_C") and the other is a description ("_C_DESCR"). This is largely done for covenience in querying the data or subsetting it without typing long strings. Because some may not be familiar with this type of data, we include a small data dictionary below to explain some of the data.

Variable                Description
---------               ---------------
Order ID                Key for order
Patient ID              Key for patient
Description             Text description of lab test
Proc Code               Procedure code for lab test
ORDER_CLASS_C_DESCR     Setting test is intended to be performed in (eg. Normal = regular blood draw)
LAB_STATUS_C            Status of laboratory result
ORDER_STATUS_C          Status of order
REASON_FOR_CANC_C       Cancellation reason (if applicable)
Order Time              Timestamp for time of original test order
Result Time             Timestamp for more recent result in the record
Review Time             Timestamp for provider acknowledgment of review of result
Department              Clinic associated with test order
ordering_route          Structure/menu in health record from which order was placed
pref_list_type          Category of preference list (if applicable)

## Data import and preparation

We have a data set that is spread out over a couple files, with varying formats for variable names, and we want to consider what data types would be most appropriate for each of our variables. The overall goal of our analysis is to understand the metadata associated with this set of orders and identify any trends that would be useful in making changes to the electronic ordering and lab or clinic workflows. At this point it might be a little abstract because we are exploring the data but we know a few things we can address up front:
- there are two files whose data could probably live in one data frame
- the column names in the file have variable formatting
- there are timestamps for which we may want to provide trends over time

**Exercise 1**

Let's work on addressing the above issues.
1. Import the data from each file
1. Clean variable names
1. Assess the relationship between the data in both files. Evaluate whether there is a one-to-one mapping, a many-to-one mapping, etc. (Hint: doing some exploration with various join functions can help answers quickly - [helpful reference](http://stat545.com/bit001_dplyr-cheatsheet.html))
1. Consider which variables you may want to represent as factors (eg. for quick visual summaries) and convert
1. Assess the time span for orders and consider if there are specific time periods over which you may want to aggregate orders to view trends (eg. daily, weekly, monthly, yearly). Add additional variables to parse out these date components (and save yourself some work in the future). Refer to lesson 4 and [lubridate documentation](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html).
1. Summarize the data

```{r, eval = FALSE}
orders_raw <-
order_details <-

orders <- orders_joined %>%

summary(orders)
```

**End Exercise**

## Exploration of data

Let's take a high level look at the data, with some areas to explore:

- Overall orders over time - are there any dramatic changes in volume over the time period?
- Which tests are most commonly ordered?
- What is the overall cancellation rate and has it changed over time?

General hint for upcoming exercises: review documentation on [janitor package](http://sfirke.github.io/janitor/articles/janitor.html) and/or [table function](https://www.rdocumentation.org/packages/base/versions/3.5.1/topics/table).

**Exercise 2**

1. Plot the order volume over the duration of the data set, at the level of day and week. (Keep in mind how ggplot parses time some geoms - it may be based on seconds.)

```{r, eval = FALSE}

```

2. Plot or tabulate the breakdown of test orders in the data set (using description or procedure code). Focus on the top 25 tests.

```{r, eval = FALSE}

```

3. Plot and/or tabulate cancellations over time for the data set.

```{r, eval = FALSE}

```

4. Explore whether there are specific tests that are cancelled more frequently than others. Focus on the top 25 tests.

```{r, eval = FALSE}

```

**End Exercise**

## Answering clinic-specific questions 

Based on some preliminary analysis and past knowledge, we want to dig into clinic-specific practices for ordering tests. 

**Exercise 3**

The following is a list of questions regarding clinic-specific characteristics of orders that we would like to answer:
- Which clinics order the highest volume of tests?
- Which clinics have the highest numbers and rates of test cancellation?
- Are there any clinics collecting blood at the clinic as opposed to at blood draw?
- Which clinics are using SmarSets (order sets) most extensively?
- Which clinics continue to use Provider Preference Lists, which are discouraged?

```{r, eval = FALSE}

```

**End Exercise**

## Evaluating turnaround times for result review

Unfortunately this data set is missing crucial timestamps needed to assess lab turnaround times. Assessing the time between order and result might be interesting, but there are various workflow variations that make this difficult to interpret. What is more straightforward to interpret, however, is the duration between when a test is resulted and when that result is reviewed by the repsonsible provider. We do not have provider identifiers in this data set, but we can still assess the result-to-review turnaround time by clinic.

**Exercise 4**

Develop a visualization that shows the distribution of different result-to-review intervals, separated by clinic.

```{r, eval = FALSE}

```

**End Exercise**
