---
title: 'Lesson 8: Beyond the csv -- parsing xml files'
author: "Randall Julian, Adam Zabell"
date: "9/3/2018"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Flat and Structured Files
The functions from readr *read_csv*, *read_tsv*, and *read_excel* (discussed in Lesson 5) open so called "Flat" files that are often called 'flat' because they have a simple structure of delimited rows and columns. Some raw instrument data is difficult to describe fully using a simple row and column approach. Several formats have been created to represent more structured data, and some have evolved into standards used across instrument applications. XML is an example of a sophisticated format that has been used to implement many standards and defacto standards. 

## The extensible markup language (xml)
XML is first, simply a text file which follows very specific grammaric rules. On top of this grammar, there is a semantic layer which is application specific. There are  packages for reading generic XML files such as **xml2**. For this lesson, we will focus on a more specific XML file format for representing mass spectrometry raw data. Using XML, several specifications (definition of how the semantic content for mass spectrometry data is represented using the XML grammar)

XML is a topic all to itself with whole books available on almost ever aspect of the system. In general, if you find you need to read an XML file, it's best to look for a package to read the specific type of XML file you need. XML was designed for programmers, and unless you have a very specific reason, often the most pragmatic approach to XML is to look for a way to avoid having to read it directly.

With standards like mzXML, mzData, mzML, and other mass spectrometry formats (including the older netCDF format - not an XML format), there are many groups who have contributed packages to CRAN to work with these files. It is important to find a package which works the way you need, since each package will have different capabilities and different requirements.

One of the more useful XML packages for mass spectrometry is currently **mzR** which is part of the **Bioconductor** suite of tools. Bioconductor is a giant collection of packages, and sometimes only part of the suite is needed. The Bioconductor group has provided a mechanism to install just a part of the overall collection, which we will use to install mzR:

```{r install_mzR, eval=FALSE}
## try http:// if https:// URLs are not supported
source("https://bioconductor.org/biocLite.R")
biocLite("mzR")

library(mzR)
```



```{r reading_mzXML, eval=FALSE}
# From the mzR docs: https://bioconductor.org/packages/devel/bioc/vignettes/mzR/inst/doc/mzR.html

mzXML <- openMSfile("7MIX_STD_110802_1.mzXML")

```

**Exercise 1:**  

```{r ex1, eval = FALSE}
# cd.tibble <- as_tibble( )
```

**End Exercise**


## Summary
* **xml2** for read/writing xml files
* **jsonlite** for read/writing json files
* troubleshoot the returned query result to catch
    + multiple values where a single was expected
    + the same key names are in every entry
    + the key names are always in the same order (or the conversion code robustly adjusts to handle this)
    + numbers stored as strings
    + strings stored as factors
