---
title: "MSACL Data Science 201 Textbook"
author:
- Patrick Mathias
- Shannon Haymond 
- Randall Julian
- Adam Zabell
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
---

# Adopting principles of reproducible research

```{r lesson_1, child = '01 - Principles of Reproducible Research.Rmd'}
```

# Getting cozy with R Markdown

```{r lesson_2, child = '02 - R Markdown.Rmd'}
```

# Reading files - beyond the basics

```{r lesson_3, child = '03 - Reading Files.Rmd'}
```

# Data manipulation in the tidyverse

```{r lesson_4, child = '04 - Data Manipulation.Rmd'}
```

# Blending data from multiple files and sources

```{r lesson_5, child = '05 - Blending Data.Rmd'}
```

# Stronger visualizations with ggplot2

```{r lesson_6, child = '06 - Visualizations.Rmd'}
```

# Exploring lab order data

```{r lesson_7, child = '07 - Exploring Data.Rmd'}
```

# Predictions using linear regression

```{r lesson_8, child = '08 - Predictions using Linear Regression.Rmd'}
```

# Classifications using linear regression

```{r lesson_9, child = '09 - Classifications using Linear Regression.Rmd'}
```

# Bringing it all together

```{r lesson_10, child = '10 - Bringing It All Together.Rmd'}
```