---
title: "Getting cozy with R Markdown"
author: "Patrick Mathias"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Why integrate your analysis and documentation in one place?

- Principle of literate programming: code is written to align with the programmer's flow of thinking
- Communication of code to others
- That communication is expected to produce better programming

## Basics of knitr and rmarkdown

- [knitr package](https://yihui.name/knitr/) and [rmarkdown package](https://cran.rstudio.com/web/packages/rmarkdown/index.html) make it easier to generate reports with integrated R code
- Integration with RStudio
- 3 components of a typical R Markdown document:
  - header
  - text
  - code chunks

## Header

```{r, eval = FALSE}
---
title: 'Lesson 2: Getting cozy with R Markdown'
author: "Patrick Mathias"
output: html_document
---
```

## Text

[R Markdown syntax](http://rmarkdown.rstudio.com/authoring_basics.html)

Examples:

- Asterisks can be used to *italicize* (`*italicize*`) or **bold** (`**bold**`) text
- Hyphens can be used to create bullet points

```{r, eval = FALSE}
- point 1
- point 2
- point 3
```

## Code chunks

    ```{r}
    mean(c(10,20,30))
    ```
    
## Flexibility in reporting: types of knitr output

3 most common output formats: 

- .html: open in most modern web browsers
- .pdf: professional looking reports
- Microsoft Word .docx: for collaborations with non-R users

Can also create variety of different presentation formats (these Powerpoint slides were generated by RStudio)

## Exercise 1

Generate a sample R Markdown document

## Exercise 2: Create output

Generate different document formats

## Exercise 3: Embed image into document

Practice embedding the "git_basic_workflow.png" diagram from the assets folder 

## Notebooks and promise/problems

- Notebooks are very helpful for interactive data analysis
- But also a problem:
  - What happens if you run code chunks out of order?
  - What happens if your environment already has existing variables with the same name?
  - Code can produce different output
- Workflow should require periodically knitting entire document and verifying output


## Summary

- Integrating code and documentation in one place produces clearer, more reproducible code
- RStudio provides useful built-in functionality for "knitting" documents into a variety of output formats
- R Markdown documents can be integrated within a recommended project structure to create a reproducible analysis

**Free online resource: https://bookdown.org/yihui/rmarkdown/**