---
title: "Getting cozy with R Markdown"
author: "Patrick Mathias"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Why integrate your analysis and documentation in one place?

- Principle of literate programming: code is written to align with the programmer's flow of thinking
- Communication of code to others
- That communication is expected to produce better programming

## Basics of knitr and rmarkdown

- [knitr package](https://yihui.name/knitr/) and [rmarkdown package](https://cran.rstudio.com/web/packages/rmarkdown/index.html) make it easier to generate reports with integrated R code
- Integration with RStudio
- 3 components of a typical R Markdown document:
  - header
  - text
  - code chunks

## Header

```{r, eval = FALSE}
---
title: 'Lesson 2: Getting cozy with R Markdown'
author: "Patrick Mathias"
output: html_document
---
```

## Text

[R Markdown syntax](http://rmarkdown.rstudio.com/authoring_basics.html)

Examples:

- Asterisks can be used to *italicize* (`*italicize*`) or **bold** (`**bold**`) text
- Hyphens can be used to create bullet points

```{r, eval = FALSE}
- point 1
- point 2
- point 3
```

## Exercise 1

1. Add a file by selecting the add file button on the top left of your screen
2. Select R Notebook as the file type
3. Find the header. Change the title of the notebook to "My First R Notebook"
4. Add your name as the author by adding another line to the header: `author: "Your Name"`
5. Add a second level heading (##) at the end of the notebook called “My Calculation”

## Code chunks

    ```{r}
    mean(c(10,20,30))
    ```

## Exercise 2: Create output

1. Within the notebook you created in Exercise 1, find the grey code chunk. Press the green play button on the top right of the chunk. What happened? Alternately, if you cursor is within a code chunk, you can hit `Ctrl+Shift+Enter`/`Command+Shift+Enter` and the code inside the chunk will execute.
2. Insert a code chunk under the cars code chunk by using the `Ctrl+Alt+i`/`Command+Option+i`(Windows/Mac) shortcut. Another option for adding a code chunk is to use the Add Code Chunk button on the top right of the Editor window (green button with a plus sign and a C).
3. Using the new code chunk you created, display the first lines of the cars data frame with the `head(cars)` command and execute the code chunk

## Flexibility in reporting: types of knitr output

3 most common output formats: 

- .html: open in most modern web browsers
- .pdf: professional looking reports
- Microsoft Word .docx: for collaborations with non-R users

Can also create variety of different presentation formats (these Powerpoint slides were generated by RStudio)

## Exercise 3: Knit other output types

1. Click the **Knit** button
2. You are being prompted to save the .Rmd file. Choose the "src" folder of your project and name the file sample_markdown_document
3. RStudio should produce output in .html format and display
4. Click the Open in Browser window and the same output should open in your default internet browser
5. If you find the folder you saved the .Rmd file there should also be a .html file you can open as well
6. Now, instead of hitting the **Knit** button, select the down arrow adjacent to it and click Knit to Word

## Adding images

- Images can be added to text sections of R Markdown document
- Important to direct R to the correct path (directory) for the image
- Syntax for incorporating an image: `![text for image caption](folder_name/image_file.ext)`

## Exercise 4

- Within the course repository that your forked and opened as an RStudio project, open up the lesson 3 file: "03 - R Markdown.Rmd"
- Embed the "git_basic_workflow.png" diagram from the "assets" folder into the lesson 3 document

## Notebooks and promise/problems

- Notebooks are very helpful for interactive data analysis
- But also a problem:
  - What happens if you run code chunks out of order?
  - What happens if your environment already has existing variables with the same name?
  - Code can produce different output
- Workflow should require periodically knitting entire document and verifying output


## Summary

- Integrating code and documentation in one place produces clearer, more reproducible code
- RStudio provides useful built-in functionality for "knitting" documents into a variety of output formats
- R Markdown documents can be integrated within a recommended project structure to create a reproducible analysis

**Free online resource: https://bookdown.org/yihui/rmarkdown/**