---
title: 'Lesson 2: Getting cozy with R Markdown'
author: "Patrick Mathias"
output: html_document
---

## Why integrate your analysis and documentation in one place?

The short answer is that it will be easier for you to understand what you did and easier for anyone else to understand what you did when you analyzed your data. This aligns nicely with the principles of reproducible research and is arguably just as important for any analysis that occurs in a clinical laboratory for operational or test validation purposes. The analysis and the explanation of the analysis live in one place so if you or someone else signs off on the work, what was done is very clear.

The more philosophical answer to this question lies in the principles of [literate programming](https://en.wikipedia.org/wiki/Literate_programming), where code is written to align with the programmer's flow of thinking. This is expected to produce better code because the program is considering and writing out logic while they are writing the code. So the advantages lie in both communication of code to others, and that communication is expected to produce better programming (analysis of data in our case).

There is another advantage of using this framework with the tools we discuss below: the output that you generate from your analysis can be very flexible. You can choose to show others the code you ran for the analysis or you can show them only text, figures, and tables. You can produce a webpage, a pdf, a Word document, or even a set of slides from the same analysis or chunks of code.

## Basics of knitr and rmarkdown

The theme of the course so far is "there's a package for that!" and this of course is no exception. The [knitr package](https://yihui.name/knitr/) and closely related [rmarkdown package](https://cran.rstudio.com/web/packages/rmarkdown/index.html) were built to make it easier for users to generate reports with integrated R code. The package documentation is very detailed but the good news is that RStudio inherently utilizes [knitr and rmarkdown](http://rmarkdown.rstudio.com/authoring_quick_tour.html) to "knit" documents and allows for a simple, streamlined workflow to create these documents.

There are 3 components of a typical R Markdown document:
- header
- text
- code chunks

### Header

The header includes metadata about the document that can help populate useful information such as title and author. This information is included in a YAML (originally *Yet Another Markup Language*, now *YAML Ain't Markup Language*) format that is pretty easy to read. For example, the header for this document is:
```{r, eval = FALSE}
---
title: 'Lesson 2: Getting cozy with R Markdown'
author: "Patrick Mathias"
output: html_document
---
```

The output field dictates the output once the document is knit, and users can add other data such as the date or even [parameters](http://rmarkdown.rstudio.com/lesson-6.html) for a report.

### Text

Text is written in whitespace sections using [R Markdown syntax](), which is a variant of a simple formatting language called markdown that makes it easy to format text using a plain text syntax. For example, asterisks can be used to *italicize* (`*italicize*`) or **bold** (`**bold**`) text and hyphens can be used to create bullet points:
- point 1
- point 2
- point 3
```{r, eval = FALSE}
- point 1
- point 2
- point 3
```


### Code chunks

Interspersed within your text you can integrate "chunks" of R code, and each code chunk can be named. You can supply certain parameters to instruct R what to do with each code chunk. The formatting used to separate a code chunk from text uses a rarely utilized character called the backtick ` that typically can be found on the very top left of your keyboard. The formatting for a code chunk includes 3 backticks to open or close a chunk and curly brackets with the opening backticks to supply information about the chunk. Here I show the general formatting in quotes:  
  

    ```{r}
    mean(c(10,20,30))
    ```
  
And if I actually write those lines with the backticks you see the code separated out:
```{r}
mean(c(10,20,30))
```

In addition code can be integrated within text by using a single backtick to open and close the integrated code. Surrounding "mean(c(10,20,30))" with backticks gives you the result in line: `mean(c(10,20,30))`.


## Flexibility in reporting: types of knitr output

## Creating an R Markdown document that integrates best practices in reproducible research
