---
title: "L2_bestPracticesExercise"
author: "Adam Zabell"
date: "1/21/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## What This Is
There is an exercise in Lesson 2 which aims to isolate the downloaded git repo by creating and following the Rstudio best practices which were outlined in Lesson 1. It becomes very easy to get turned around, especially if you spent most of Sunday downloading (and re-downloading) files while also following along with the conversation.

Let's walk through each of the steps in that exercise, which includes the point where we build this file. 

### 1. Close the RStudio window that has your "sample-project-structure" project open
Lesson 1 showed how to create a sample project, and we used it for the rest of that lesson and the first half of Lesson 2. We're just closing it before moving back to "our main project."

### 2. Select your "MSACL-intermediate-R-course" project
This is that main project, which you built when you created the project which connected to the git repository for all the class lessons. There should be ten folders (lesson1 .. lesson10) and a handful of R markdown, R script, and html files.

### 3. Create a new directory called "coursework" and set your working directory
In order to follow best practices for any project you're working on, you should be using the basic project structure of 'each file type in the right subfolder' which is the right practice in general but it clashes with the repo structure of 'one lesson per subfolder.' The solution we've chosen to follow will build that structure within the project. And because we're within a project, the current working directory we're starting from should be the initial project directory. As a result, we can use relative directory shorthand when resetting the working directory.  
```{r new_directory}
dir.create("coursework")
setwd("./coursework")
```

### 4. Load the here package and use `set_here()` 
This seems to be the most tricky part of the process. Rather than belabor the issue, we'll simply ensure that our current working directory is what we want, then load the library for `here`, and confirm that the variable is set like we want. In any future R markdown files from this point onward, that library load should happen in the header of the file. For clarity, we waited to add it until now.
```{r set_up_the_here}
getwd() #should be /some/diretory/path/then/git-repository-name/coursework
library(here)
```

### 5. Restart the **R** session
This clears anything weird which may be hanging around in your session, including some points which could complicate the `here` package. Step four created the ".here" file in the "coursework" directory, so you're fine there. However, restarting the session means, after the restart, you'll need to read in the library again. 
```{r restart_and_relibrary}
# Restarting R is a dropdown command in Rstudio and
# can not be done from within a code block. But 
# once you've done that, you should execute this next line
library(here)
```

### 6. Create the best-practices directory structure
We won't necessarily use them as we step through the lessons. But when you get to the final exercise of Lesson 10 it would be wise to take advantage of this structured layout.
```{r make_some_dirs}
dir.create("data") # raw data
dir.create("output") # output from analysis
dir.create("cache") # intermediate data (after processing raw data)
dir.create("src") # code goes into this folder
```

### 7. Copy the "class_data" directory into the "coursework" directory
It's probably easiest to do this using your computer's operating system. However, assuming you put "class_data" where we told you to, the following code block will accomplish the same task. Strictly speaking we should be copying the files from "class_data" into the folder we just created called "data" but our lessons were written assuming this directory name and we'd rather not edit them at this point.
```{r copy_class_data}
file.copy(
  from = here::here("..","class_data"),
  to = here::here(),
  recursive = TRUE
)
```

### 8. Copy the lesson Rmd files into the "src" directory
It would certainly be good to sync your local repository to the master repository on git before doing this, since there's a good chance the lessons have been updated since yesterday. Once that is done, you should copy them into the source code folder so that you can work with them without disturbing the originals.
```{r copy_rmd}
# Synching to the git repo is not something you can do
# from within a code block, but once you have, execute
# the following
lesson_dirs <- c("lesson1","lesson2","lesson3","lesson4","lesson5",
                 "lesson6","lesson7","lesson8","lesson9","lesson10")
for(l in lesson_dirs) {
  rmd_file <- paste0(l,".Rmd")
  file.copy(
    from = here::here("..",l,rmd_file),
    to = here::here("src")
  )
}
```

### N. Write and then save this file 
So techincally this file isn't discussed in Lesson 2. But you would have been following these steps and entering them in either an R script file, an R markdown file, or directly into the console. If this is work you did in the project, best practices says you should document what you did, and R markdown is often the best way to do that. So let's save all of these steps in a markdown file, and then save the file within the "src" directory. This is at least the third time we need to use the Rstudio GUI and can't just run a code block to do an action. Click on "File .. Save .." and save this file as *meta_lesson.Rmd*

On the other hand, if you're pulling this file down from the git repository, it already exists within the folder for Lesson 2.